Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\tnoteref{t1,t2}

### 2016-04-28 curvature.tex
reference tweaks

\subsection{Simulating random walks on the rSPR graph.}
\label{sec:random_walks}
The uniform random walk moves from one vertex to one of its neighbors uniformly at random, which makes this walk more likely to sample higher degree vertices.
In contrast, the Metropolis Hastings (MH) random walk with constant likelihood function proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
The MH random walk is guaranteed to sample each tree uniformly at random and is therefore representative of a phylogenetic MCMC program sampling trees under a uniform prior.

@@

Now, we show that the average size of the moved subtrees is small.

\begin{lem}
	Let $T$ be a rooted tree with $n \ge 4$ leaves.
	Let $N_i(T)$ be the neighbors of $T$ induced by rSPR moves of an $i$-leaf subtree.
	Then $$\frac{\sum_i i \size{N_i(T)}}{\size{N(T)}} \le \frac{n + 3}{4} .$$
	\label{lem:spr_size_mean}
\end{lem}
\begin{proof}
	Again recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
	By similar arguments as in the proof of Lemma~\ref{lem:spr_size}, a ladder tree has the greatest average moved subtree size (in terms of the number of leaves).
	There are $n$ subtrees with 1 leaf, and 1 subtree each with $2,3, \ldots n - 1$ leaves.
	We ignore the root subtree, which does not induce any rSPR moves.
	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a 1-leaf subtree.
	We can bound the distribution of the remaining moves with the (loose) assumption that each other subtree contributes the same number of moves.
	In particular, we assume that each subtree contributes $2n - 3 - 5$ moves.
	This underestimates the number of small moved subtrees and overestimates the number of larger moved subtrees.
	Thus, this estimate bounds the mean moved subtree size from above.
	The mean subtree size, ignoring small subtrees, is $(3 + (n-2)) / 2 = (n + 1) / 2$.
	Thus, the average size of a moved subtree over all SPR operations from a given tree is at most the weighted average of $1 / 2 + ((((n + 1) / 2) / 2$ = $(n + 1) / 4 + 1 / 2 = (n + 3) / 4$.

>>>>

\subsection{Simulating random walks on the rSPR graph.}
\label{sec:random_walks}
The uniform random walk moves from a vertex to one of its neighbors uniformly at random, which makes this walk more likely to sample higher degree vertices.
In contrast, the Metropolis Hastings (MH) random walk with constant likelihood function proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
The MH random walk is guaranteed to sample each tree uniformly at random and is therefore representative of a phylogenetic MCMC program sampling trees under a uniform prior.

@@

Now, we bound the average size of the moved subtrees.

\begin{lem}
	Let $T$ be a rooted tree with $n \ge 4$ leaves.
	Let $N_i(T)$ be the neighbors of $T$ induced by rSPR moves of an $i$-leaf subtree.
	Then $$\frac{\sum_i i \size{N_i(T)}}{\size{N(T)}} \le \frac{n + 3}{4} .$$
	\label{lem:spr_size_mean}
\end{lem}
\begin{proof}
	Again recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
	By similar arguments as in the proof of Lemma~\ref{lem:spr_size}, a ladder tree has the greatest average moved subtree size (in terms of the number of leaves).
	There are $n$ subtrees with 1 leaf, and 1 subtree each with $2,3, \ldots n - 1$ leaves.
	We ignore the root subtree, which does not induce any rSPR moves.
	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a small subtree.
	We can bound the distribution of the remaining moves with the (loose) assumption that each other subtree contributes the same number of moves.
	In particular, we assume that each subtree contributes $2n - 3 - 5$ moves.
	This underestimates the number of small moved subtrees and overestimates the number of larger moved subtrees.
	Thus, this estimate bounds the mean moved subtree size from above.
	The mean subtree size, ignoring small subtrees, is $(3 + (n-2)) / 2 = (n + 1) / 2$.
	Thus, the average size of a moved subtree over all SPR operations from a given tree is at most the weighted average of $1 / 2 + ((((n + 1) / 2) / 2$ = $(n + 1) / 4 + 1 / 2 = (n + 3) / 4$.
